Degeneracy in a restricted Boltzmann machine

### 2016-02-17 writing/degeneracy/draft.Rmd
started on poster

#Degeneracy in Random Graph Models
In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. @handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}.

@@

@schweinberger2011instability introduces a related concept, instability, which  is characterized
by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the a discrete exponential family model, and thus near-degeneracy.

@@

@@

##Small example
To illustrate the idea of graph degeneracy in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

@@

```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model. For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also."}
plot_dat_grid %>% 
  mutate(dist = sqrt(r1^2 + r2^2)) %>%
  group_by(dist, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n()) %>%
  mutate(Hiddens = as.character(H), Visibles = paste0("Visibles: ", V)) %>%
  ggplot() +
  geom_point(aes(dist, frac_degen, colour = Hiddens)) +
  geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
  facet_grid(~Visibles) +
  ylim(c(0,1)) +
  xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
  ylab("Fraction Near-degenerate")
```

>>>>

#Degeneracy in Random Graph Models
In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. We consider three related definitions of degeneracy as they relate to a graphical model. Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between marginal and conditional expectations of the model.

##Mean Parameterization
@handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}.

@@

##Neg-potential
@schweinberger2011instability introduces a related concept, instability, which  is characterized
by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the a discrete exponential family model, and thus near-degeneracy.

@@

##Conditional Expectation

@@

#Tiny example
To illustrate the idea of graph degeneracy in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

##Degeneracy three ways

@@

```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model. For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=3}
plot_dat_grid %>% 
  mutate(dist = sqrt(r1^2 + r2^2)) %>%
  group_by(dist, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n()) %>%
  mutate(Hiddens = as.character(H), Visibles = paste0("Visibles: ", V)) %>%
  ggplot() +
  geom_point(aes(dist, frac_degen, colour = Hiddens)) +
  geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
  facet_grid(~Visibles) +
  ylim(c(0,1)) +
  xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
  ylab("Fraction Near-degenerate")
```
